Evolution of precopulatory and postcopulatory inbreeding behaviour

### 2016-08-25 ms_pre_post.Rmd
Add Density regulation section, Model analysis section, and a mutation section (inserted earlier) into the Model

---
title: "Evolution of precopulatory and postcopulatory inbreeding behaviour"
author: "A. Bradley Duthie et al."
date: "August 24, 2016"
output:
  pdf_document:
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
linestretch: 1
bibliography: ms_pre_post.bib
---

@@

Values of phenotypic traits for tendency to engage in polyandry ($P_{p}$), mating ($M_{p}$), and fertilistation ($F_{p}$) each have costs, which independently increase the probability of a focal female's reproductive failure. These costs are all expressed prior to mating. In the case of tendency to engage in polyandry and inbreeding mating strategy, costs of $P_{p}$ ($c_{P}$) and $M_{p}$ ($c_{M}$) are interpreted biologically as costs associated with mate searching. In the case of inbreeding fertilisation strategy, the cost of $F_{g}$ is interpreted as an energetic cost associated with the physiological or biochemical mechanisms that need to be developed to bias fertilsation for inbreeding avoidance or inbreeding preference. Each cost independently increases the risk of female pre-mating mortality and, because generations are non-overlapping, total reproductive failure as a linear function of its associated trait. The probability that a focal female dies before mating due to the cost of polyandry is therefore $P_{p} \times c_{P}$. Similarly, the probability of pre-mating female mortality due to the cost of mating strategy and fertilisation strategy is $|M_{p}| \times c_{M}$ and $|F_{p}| \times c_{F}$, repectively, where $|M_{p}|$ is the absolute value of $M_{p}$ and $|F_{p}|$ is the absolute value of $F_{p}$. Absolute values are used for applying costs to inbreeding strategies because both negative and positive $M_{p}$ and $F_{p}$ bias breeding interactions with respect to kinship, while only positive values of $P_{p}$ affect a female's tendency for polyandry; zero and negative $P_{p}$ values always translate to monandrous phenotypes (see below).

@@

@@

The viability of a focal female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a function of the kinship coefficient between $i$ and the sire of her offspring $j$ ($k_{ij}$), and a log-linear slope of inbreeding depression $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
The slope $\beta$ models the number of haploid lethal equivalents that exist in the gametes of $i$ and $j$, and which might combine to become homozygous in offspring and reduce viability. Equation 1 assumes that each deleterious recessive allele affects offspring viability independently, meaning that every deleterious recessive allele added will have a multiplicative affect on decreasing viability [@Morton1956; @Mills1994], causing the log-linear relationship between parent kinship and inbreeding depression. We assume that inbreeding depression always decreases offspring viability, $\beta \geq 0$ (i.e., no outbreeding depression). Therefore, because $k_{ij}$ is constrained to values between zero and one, the product of $-\beta$ and $k_{ij}$ must be a value less than or equal to zero, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero and one. We can therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability using a Bernoulli trial.

@@

>>>>

---
title: "Evolution of precopulatory and postcopulatory inbreeding behaviour"
author: "A. Bradley Duthie et al."
date: "August 24, 2016"
output:
  pdf_document:
    fig_caption: yes
  html_document: default
  word_document: null
linestretch: 1
bibliography: ms_pre_post.bib
---

@@

Values of phenotypic traits for tendency to engage in polyandry ($P_{p}$), mating ($M_{p}$), and fertilistation ($F_{p}$) each have costs, which independently increase the probability of a focal female's reproductive failure. These costs are all expressed prior to mating. In the case of tendency to engage in polyandry and inbreeding mating strategy, costs of $P_{p}$ ($c_{P}$) and $M_{p}$ ($c_{M}$) are interpreted biologically as costs associated with mate searching or courtship [e.g., @Rowe1988; @Rowe1994; @Ronkainen1994; @Koga1998; @Lasley-Rasher2012]. In the case of inbreeding fertilisation strategy, the cost of $F_{p}$ ($c_{F}$) is interpreted as an energetic cost associated with the physiological or biochemical mechanisms that need to be developed to bias fertilsation for inbreeding avoidance or inbreeding preference [e.g., @Gasparini2011]. Each cost independently increases the risk of female pre-mating mortality and, because generations are non-overlapping, total reproductive failure as a linear function of its associated trait. The probability that a focal female dies before mating due to the cost of polyandry is therefore $P_{p} \times c_{P}$. Similarly, the probability of pre-mating female mortality due to the cost of mating strategy and fertilisation strategy is $|M_{p}| \times c_{M}$ and $|F_{p}| \times c_{F}$, repectively, where $|M_{p}|$ is the absolute value of $M_{p}$ and $|F_{p}|$ is the absolute value of $F_{p}$. Absolute values are used for applying costs to inbreeding strategies because both negative and positive $M_{p}$ and $F_{p}$ bias breeding interactions with respect to kinship, while only positive values of $P_{p}$ affect a female's tendency for polyandry; zero and negative $P_{p}$ values always translate to monandrous phenotypes (see below).

@@

Mutation
----------------------------------------------

Alleles of a focal female's offspring mutate with a probability of $\mu=0.001$, with mutation being an independent event across loci. In general, loss of function mutations might be expected to be more frequent than mutations that increase the functionality of adaptations that rely on complex physiologies or behaviours. If tendency for polyandry, inbreeding mating strategies, or inbreeding fertilisation strategies rely on such complex physiologies or behaviours, then we might expect the direction of mutation to be biased (e.g., towards monandry or random mating). However, because we are interested specifically in how selection affects reproductive strategy given varying costs and inbreeding depression, we model mutation as unbiased so that selection can be isolated and inferred clearly from mean allele values.

Hence, when a mutation occurs at a loci, a new value is sampled from a random normal distribution with a mean of zero, $\mathcal{N}(0, \mu_{SD})$. This new value is then added to the original value of the allele that mutates [@Kimura1965; @Lande1976; @Bocedi2014; @Duthie]. The standard deviation of the mutation effect size ($\mu_{SD}$) is set to $\sqrt{1/20}$, which scales the variance of mutation effect size to the total number of alleles affecting $P_{p}$, $M_{p}$, and $F_{p}$ phenotypes [see @Kimura1965; @Lande1976; @Bocedi2014].

@@

The viability of a focal female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a function of the kinship coefficient between $i$ and the sire of her offspring $j$ ($k_{ij}$), and a log-linear slope of inbreeding depression $\beta$. We model selection as having an absolute effect on offspring viability (i.e., hard selection), rather than a relative effect on viability (i.e., soft selection), so that the effect of $\beta$ is consistent over generations and across different parameter combinations. Natural selection on tendency for polyandry ($P_{p}$), inbreeding mating strategy ($M_{p}$), and inbreeding fertilisation strategy ($F_{p}$) can therefore be compared with respect to log-linear inbreeding depression slopes across generations and parameter combinations. If offspring viability was instead relative such that the least inbred offspring had a high viability regardless of the absolute value of $k_{ij}$, then the effect of kinship coefficients and inbreeding depression slopes on offspring viability would depend on the population-wide distribution of $k_{ij}$ between females and males, making model inference more difficult. Therefore, we define offspring viability as,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
The slope $\beta$ models the number of haploid lethal equivalents that exist in the gametes of $i$ and $j$, and which might combine to become homozygous in offspring and reduce viability. Equation 1 assumes that each deleterious recessive allele affects offspring viability independently, meaning that every deleterious recessive allele added will have a multiplicative affect on decreasing viability [@Morton1956; @Mills1994], causing the log-linear relationship between parent kinship and inbreeding depression. We assume that inbreeding depression always decreases offspring viability, $\beta \geq 0$ (i.e., no outbreeding depression). Therefore, because $k_{ij}$ is constrained to values between zero and one, the product of $-\beta$ and $k_{ij}$ must be a value less than or equal to zero, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero and one. We can therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability using a Bernoulli trial.

@@

To avoid indefinite population growth over generations, we introduce a separate carrying capacity on the total number of females ($K_{f}$) and males ($K_{m}$) in the focal population following immigration [@Guillaume2009; @Duthie]. If at the end of a generation the number of females or males exceeds their respective carrying capacities, then each sex is reduced back to its $K_{f}$ or $K_{m}$ by randomly removing females and males until each sex is at its carrying capacity. Removal of females and males can be interpreted as dispersal outside the focal population, mortality, or as some combination of dispersal and mortality. The remaining females and males within the population form the next generation of adults.

Model analysis
----------------------------------------------

To analyse our model, we track mean values of alleles underlying tendency for polyandry ($P_{a}$), inbreeding mating strategy ($M_{a}$), and inbreeding fertilisation strategy ($F_{a}$) in a focal population. We present evolutionary dynamics of these mean allele values across a range of costs and inbreeding depression slopes using two types of simulations. First, we consider eight possible cost parameter combinations in which costs of tendency for polyandry ($c_{P}$), inbreeding mating strategy ($c_{M}$), and inbreeding fertilisation strategy ($c_{F}$) equal either $0$ or $0.02$ [@Duthie observed $0.02$ to be strong cost of tendency for polyandry]. Second, we fix $c_{P}$ to values of $0$, $0.01$, or $0.02$, then randomly select $c_{M}$ and $c_{F}$ values between $0-0.05$ for simulation initial starting conditions. The objective of the first type of simulation is to test how mean allele values change over generations in response to different combinations of costs and inbreeding depression slopes. The objective the second type of simulation is to predict selection for tendency for polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy, across a wider range of parameter combinations, and to therefore test under what conditions each is predicted to evolve given different combinations of costs and inbreeding depression slopes.

In the first type of simulation, we record dynamics of mean values of $P_{a}$, $M_{a}$, $F_{a}$ over generations for every possible combination in which the cost associated with each of the three phenotypes ($c_{P}$, $c_{M}$, and $c_{F}$) is either $0$ or $0.02$, and for four slopes of inbreeding depression on offspring viability $\beta = {0, 1, 2, 3}$. Each possible cost combination is replicated ten times, and grand mean values and standard deviation of means is calculated in each generation across replicate simulations. These analyses allowed us to infer how allele values change over generations in response to costs, but also in response to the changing values of other alleles and therefore potentially interacting phenotypes (e.g., selection on inbreeding fertilisation strategy might be weak if polyandry is rare).

In the second type of simulation, we record and present mean values of $P_{a}$, $M_{a}$, $F_{a}$ only from the terminal generation of simulations, and for random costs of inbreeding mating strategy $c_{M} \in [0, 0.05]$ and inbreeding fertilisation strategy $c_{F} \in [0, 0.05]$, but fixed costs of tendency for polyandry $c_{P} = {0, 0.01, 0.02}$. We consider four values of $\beta = {0, 1, 2, 3}$, and we simulate 100 replicates for all combinations of $c_{P}$ and $\beta$. For different values of $\beta$ given the same $c_{P}$ value, we use the same randomly selected values of $c_{M}$ and $c_{F}$ to better infer how terminal mean allele values are affected by different magnitudes of inbreeding depression. We do not use the same randomly selected values of $c_{M}$ and $c_{F}$ across different values of $c_{P}$ because $c_{P}$ values were fixed *a priori* only to facilitate model inference. The objective of randomising costs was to sample from a broad range of parameter space, and this range is increased by selecting different $c_{M}$ and $c_{F}$ values for each $c_{P}$.

For each replicate simulation with a unique set of costs, we classify the evolutionary outcome of inbreeding stragies in the simulation in one of four ways: (1) Weak selection for inbreding mating strategy and inbreeding fertilisation strategy, (2) weak selection for inbreeding mating strategy and strong selection for inbreeding fertilisation strategy, (3) strong selection for inbreeding mating strategy and weak selection for inbreeding fertilisation strategy, and (4) strong selection for inbreeding mating strategy and inbreeding fertilisation strategy. We additionally classify the evolutionary outcome of tendency for polyandry in the simulation as either weak selection for polyandry or strong selection for polyandry. @Duthie showed that selection for inbreeding preference (positive values of $M_{a}$) is predicted when inbreeding depression on offspring viability is $\beta=0$, but selection for inbreeding avoidance (negative values of $M_{a}$) is predicted when $\beta \geq 1$. When referring to strong selection for inbreeding mating strategy and inbreeding fertilisation strategy, we therefore refer specifically to inbreeding preference when $\beta=0$ and inbreeding avoidance when $\beta \geq 1$. Strong selection is inferred when mean allele values exceed $1$ (when $\beta = 0$) or are less than $-1$ (when $\beta \geq 1$), otherwise selection is inferred to be weak. Exploratory simulations suggest that cutoff values of $1$ and $-1$ are conservative for inferring when selection is strong, and other cutoff values do not change the general pattern of evolutionary outcomes (1-4) across cost parameter values.

For all replicates in both the first and second simulation types, we set the maximum number of generations to 5000, which exploratory simulations and previous modelling [@Duthie] showed to be a sufficient number of generations for inferring long-term dynamics of mean allele values and therefore selection on phenotypes. For all replicates, we set the number of immigrants per generation to $\rho=5$, which produced a range of kin and non-kin in each generation such that females could express inbreeding strategies. Females always produced $n=8$ offspring, which was sufficient to keep populations consistently near carrying capacities and avoid population extinction. Carrying capacities for females ($K_{f}$) and males ($K_{m}$) were both set to 100 because previous modelling has shown that populations of this size are small enough that mating encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but populations are also not so small that selection is typically overwhelmed by genetic drift (Duthie and Reid *in press*).
